---
title: "trajectory_analysis"
author: "Maria Firulyova"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r load_libs}
suppressMessages(library(babelwhale))
suppressMessages(library(dyno))
suppressMessages(library(dplyr))
suppressMessages(library(tidyverse))
suppressMessages(library(sctransform))
suppressMessages(library(SeuratData))
suppressMessages(library(Seurat))
```

## Info

* Article: https://www.nature.com/articles/s41587-019-0071-9
* Quick start: https://dynverse.org/users/2-quick_start/
* Methods: https://dynverse.org/reference/dynmethods/

## Requirements

* singularity (version > 3): https://anaconda.org/conda-forge/singularity
* dyno: https://dynverse.org/users/1-installation/
* Seurat

## Config singularity

Dyno authos wrap multiple methods for trajectory inference into both Docker and Singularity containers.
Since docker requires sudo, I use singularity.

```{r}
set_default_config(create_singularity_config(cache_dir = 'sin_cahce'))
```

## Define some functions

```{r}
get_data <- function(obj, assay = 'SCT') {
  wrap_data <- wrap_expression(counts = t(as.matrix(obj@assays[[assay]]@counts)),
                               expression = t(as.matrix(obj@assays[[assay]]@data)))
  wrap_data <- add_grouping(wrap_data,
                            obj$seurat_clusters)
  wrap_data
}

get_branching_point <- function(model, obj) {
  branching_milestone <- model$milestone_network %>%
    group_by(from) %>%
    filter(n() > 1) %>%
    pull(from) %>% dplyr::first()
  branch_point <-
    calculate_branching_point_feature_importance(model,
                                                 expression_source =
                                                   obj$expression,
                                                 milestones_oi = branching_milestone)
  branch_point
}

```


## Prepare Seurat object

```{r}
set.seed(1)

get_seurat <- function(obj) {
  data("pbmc3k")
  pbmc3k <- pbmc3k[, sample(colnames(pbmc3k), size = 5e2, replace = F)]
  pbmc3k <-
    PercentageFeatureSet(pbmc3k, pattern = "^MT-", col.name = "percent.mt")
  pbmc3k <-
    SCTransform(pbmc3k, vars.to.regress = "percent.mt", verbose = FALSE)
  pbmc3k <- RunPCA(pbmc3k, verbose = FALSE)
  pbmc3k <- RunUMAP(pbmc3k, dims = 1:30, verbose = FALSE)
  pbmc3k <- FindNeighbors(pbmc3k, dims = 1:30, verbose = FALSE)
  pbmc3k <- FindClusters(pbmc3k, verbose = FALSE)
  pbmc3k
}

pbmc3k <- get_seurat()
```

## Convert seurat object to input data for dyno wrapper

```{r}
obj <- get_data(pbmc3k)
str(obj)
```

## Infer thajectory

For demonstration purpose I choose the slingshot method, however you can run multiple methods by one command (see `?infer_trajectory`).

```{r}
set.seed(1)
model <- infer_trajectory(obj, method = 'slingshot', verbose = TRUE)
```

## Get UMAP for better visualization

```{r}
dimred <- dyndimred::dimred_umap(obj$expression)

```

## Calculate some features: feature importance, branch points

```{r}
branch_feature_importance <-
  calculate_branch_feature_importance(model, expression_source = obj$expression)
branch_feature_importance
```

```{r}
branch_point <- get_branching_point(model, obj)
branch_point
```

## Visualize trajectory

```{r}
plot_dimred(
  model,
  dimred = dimred,
  expression_source = obj$expression,
  color_density = "grouping",
  alpha_cells = 0.7,
  grouping = obj$grouping,
  label_milestones = FALSE,
) +
  theme(aspect.ratio = 1, legend.position = "none")
```

## Genes important at bifurcation points

```{r}
plot_heatmap(
  model,
  expression_source = obj,
  features_oi = unique(branch_point %>% top_n(30, importance) %>% pull(feature_id))
)
```

## Lineage / branch markers: "3" lineage

```{r}
features <- branch_feature_importance %>%
  filter(to == "3") %>%
  top_n(30, importance) %>%
  pull(feature_id)
plot_heatmap(model, expression_source = obj, features_oi = features) +
  ggtitle('features important for branch to 3')
```